Synthesizing Quotient Lenses

### 2018-06-24 papers/QuotientLenses/main.tex
Fixed some typos

\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs
(Optician), 
and we have a way of defining regular expressions with equivalence
relations indicating the essential information (QREs).
A tantalyzing possibility would be to use the bijective string lens
synthesis procedure as a subroutine for a quotient lens synthesis procedure. 
This new synthesizer would take as input source and target QREs and 
example input-output pairs, 
compute the canonical source/target formats from the QREs, 
map the example input-output pairs to their canonical representations,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the canonical examples.

@@

\noindent
Intuitivley, $W(Q)$ is the regular expression representing the
external format, while $K(Q)$ is the regular expression representing
the internal format.  The equivalence relation $\eqrel{Q}$ groups
together elements in the language of $W(Q)$ that contain the same
essential information.  The function $\canonize(Q)$ picks the
representative element from each of the resulting equivalence classes.

@@

To assess the performance of QRE synthesis, we are interested in two different
questions. First, how does the performance of \QOpt{} compare to the performance
of Optician on benchmarks that do not require QREs? The answer to this question
tells us how much overhead we have introduced by adopting the more general
mechanism. Figure~\ref{fig:times}(a) shows that \QOpt{} was able to synthesize
all of the Optician benchmarks at a speed competative with the old version.
There is a small amount of additional overhead introduced by QREs in calculating
the $W$ and $K$ functions, resulting in a slight decrease in performance.

@@

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parantheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are ambiguous, and this ambiguity can cause
unintended updates to be made the source, especially when information is
discarded in this process.

\iffalse
For example, imagine that a book database has source type \\
$S$ = \lstinline{books[book[title[string], author[string]+]*]}
with view type $V$ = \lstinline{title[string]}. When the source type is
transformed to the view type for matching, only titles are kept, the element
labels are destroyed and all authors are replaced for the empty sequence, thus
leaving the intermediate type $S'$ = \lstinline{(title[string],(),()*)*}. In
BiFluX, $V$ is considered a subtype of $S'$ because the flat values of $V$ are
also flat values of $S'$. Since the intermediate type is $S'$ ambiguous, a view
sequence \lstinline{[title["mybook"]]} would be upcast into an intermediate
value \lstinline{[(title["mybook"],((), [ ]))]} that would in turn lead to an
updated source \lstinline{books[[book[(title["mybook"],(aut, [ ]))]]]}, which is
clearly unsatisfactory as it would discard the entire book database except the
first author \lstinline{aut} of the first book.
\fi
The typechecking phase of a BiFluX program therefore includes a {\em type
normalization} phase that tries to normalize a source types using automata
reduction techniques, and derive a lens between ambiguous and unambiguous
types. On the view side, this phase only tries to normalize view types into
isomorphic types. Pacheco et. al. do not claim that their normalization
procedures are complete in the sense that they can disambiguate any ambiguous
type, but they do show that when these procedures succed the normalized types
are unambiguous.

@@

More concretely, their approach gives the developer an explicit option to
formulat her understanding of two object models being different. The technical
realization is as follows: The developer specifies ``classifying term'', a
closed Object Constraint Language (OCL) query term that can be evaluated in an
object model and returns a characteristic value. Two object models with the same
characteristic value belong to the same equivalence class. For example under a
configuration requiring at least 2 and at most 4 Person objects, the
classifying term \lstinline{Person.allInstances()->size()} would yield three
object models with 2,3 and 4 Person objects respectively.

Classifying terms are therefore analogous to our QREs or to Boomerang canonizers
in the sense that they define an equivalence relation on data. However, while
QREs and canonizers aim at identifying or characterizing the elements that
should be treated equally by a lens, the focus of the work by Hilken et. al. is
on classes that represent specific patterns of particular relevance to the
modeler who is interested in analyzing the behavior of a transformation.

@@

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences. Perhaps the system most similar to Optician is
InSynth~\cite{gvero-pldi-2013}, a system for synthesizing terms in the
simply-typed lambda calculus that addresses equivalences on types. Instead of
trying to directly synthesize terms of the simply-typed lambda calculus,
InSynth synthesizes a well-typed term in the succinct calculus, a language with
types that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, Anders et. al. used a pseudo-canonical form that
captures part of the equivalence relation over types were used. To preserve
completeness, they pushed some of the remaining parts of the type equivalence
relation into a set of rewriting rules and other parts into the synthesis
algorithm itself.

>>>>

\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs
(Optician), 
and we have a way of defining regular expressions with equivalence
relations indicating the essential information (QREs).
A tantalizing possibility would be to use the bijective string lens
synthesis procedure as a subroutine for a quotient lens synthesis procedure. 
This new synthesizer would take as input source and target QREs and 
example input-output pairs, 
compute the canonical source/target formats from the QREs, 
map the example input-output pairs to their canonical representations,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the canonical examples.

@@

\noindent
Intuitively, $W(Q)$ is the regular expression representing the
external format, while $K(Q)$ is the regular expression representing
the internal format.  The equivalence relation $\eqrel{Q}$ groups
together elements in the language of $W(Q)$ that contain the same
essential information.  The function $\canonize(Q)$ picks the
representative element from each of the resulting equivalence classes.

@@

To assess the performance of QRE synthesis, we are interested in two different
questions. First, how does the performance of \QOpt{} compare to the performance
of Optician on benchmarks that do not require QREs? The answer to this question
tells us how much overhead we have introduced by adopting the more general
mechanism. Figure~\ref{fig:times}(a) shows that \QOpt{} was able to synthesize
all of the Optician benchmarks at a speed competitive with the old version.
There is a small amount of additional overhead introduced by QREs in calculating
the $W$ and $K$ functions, resulting in a slight decrease in performance.

@@

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parentheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are ambiguous, and this ambiguity can cause
unintended updates to be made the source, especially when information is
discarded in this process.

\iffalse
For example, imagine that a book database has source type \\
$S$ = \lstinline{books[book[title[string], author[string]+]*]}
with view type $V$ = \lstinline{title[string]}. When the source type is
transformed to the view type for matching, only titles are kept, the element
labels are destroyed and all authors are replaced for the empty sequence, thus
leaving the intermediate type $S'$ = \lstinline{(title[string],(),()*)*}. In
BiFluX, $V$ is considered a subtype of $S'$ because the flat values of $V$ are
also flat values of $S'$. Since the intermediate type is $S'$ ambiguous, a view
sequence \lstinline{[title["mybook"]]} would be upcast into an intermediate
value \lstinline{[(title["mybook"],((), [ ]))]} that would in turn lead to an
updated source \lstinline{books[[book[(title["mybook"],(aut, [ ]))]]]}, which is
clearly unsatisfactory as it would discard the entire book database except the
first author \lstinline{aut} of the first book.
\fi
The typechecking phase of a BiFluX program therefore includes a {\em type
normalization} phase that tries to normalize a source types using automata
reduction techniques, and derive a lens between ambiguous and unambiguous
types. On the view side, this phase only tries to normalize view types into
isomorphic types. Pacheco et. al. do not claim that their normalization
procedures are complete in the sense that they can disambiguate any ambiguous
type, but they do show that when these procedures succeed the normalized types
are unambiguous.

@@

More concretely, their approach gives the developer an explicit option to
formulate her understanding of two object models being different. The technical
realization is as follows: The developer specifies ``classifying term'', a
closed Object Constraint Language (OCL) query term that can be evaluated in an
object model and returns a characteristic value. Two object models with the same
characteristic value belong to the same equivalence class. For example under a
configuration requiring at least 2 and at most 4 Person objects, the
classifying term \lstinline{Person.allInstances()->size()} would yield three
object models with 2,3 and 4 Person objects respectively.

Classifying terms are therefore analogous to our QREs or to Boomerang canonizers
in the sense that they define an equivalence relation on data. However, while
QREs and canonizers aim at identifying or characterizing the elements that
should be treated equally by a lens, the focus of the work by Hilken et al. is
on classes that represent specific patterns of particular relevance to the
modeler who is interested in analyzing the behavior of a transformation.

@@

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences. Perhaps the system most similar to Optician is
InSynth~\cite{gvero-pldi-2013}, a system for synthesizing terms in the
simply-typed lambda calculus that addresses equivalences on types. Instead of
trying to directly synthesize terms of the simply-typed lambda calculus,
InSynth synthesizes a well-typed term in the succinct calculus, a language with
types that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, Anders et al. used a pseudo-canonical form that
captures part of the equivalence relation over types were used. To preserve
completeness, they pushed some of the remaining parts of the type equivalence
relation into a set of rewriting rules and other parts into the synthesis
algorithm itself.
